%------------------------------------------------

\begin{fullwidth}
Most of the field research done in development involves human subjects -- real people.
As a researcher, you are asking people to trust you with personal information about themselves:
where they live, how rich they are, whether they have committed or been victims of crimes,
their names, their national identity numbers, and all sorts of other data.
Development research does not just involve real people -- it also affects real people.
Policy decisions are made every day using the results of briefs and studies,
and these can have wide-reaching consequences on the lives of millions.
As the range and importance of policy-relevant questions being asked by development researchers grow,
so too does the (rightful) scrutiny that methods and results are placed under.

Standards and transparency are important so that the quality of research can be judged on its merits.
Without a transparent measure of credibility, reputation is a primary signal, and two failures may occur:
low-quality studies from reputable sources may be used as evidence when in fact they don't warrant it,
and high-quality studies from sources without an international reputation may be ignored.
Both these outcomes reduce the quality of evidence overall,
but even more importantly they usually shift the balance of power in development research
away from the people and places who are most directly affected by it.
Simple transparency standards mean that high-quality research is more common, and more impactful.
This section will provide some basic guidelines and resources
for collecting, handling, and using field data ethically and responsibly to publish research findings.
\end{fullwidth}

\section{Personally-identifying information}

Anytime you are collecting primary data in a development research project,
you are almost certainly handling data that include \textbf{personally-identifying information} (PII).
\index{Personally-identifying information}
PII data carries strict expectations about storage and handling,
and it is the responsibility of the research team to satisfy these expectations.
Your funder or employer will most likely require you to hold a certification from a source
such as Protecting Human Research Participants\sidenote{\url{https://phrptraining.com/}}
or the CITI Program\sidenote{\url{https://about.citiprogram.org/en/series/human-subjects-research-hsr/}}.

PII data contains information that can, without any transformation, be used to identify
individual people, households, villages, or firms that were included in data collection.
This includes names, addresses, and geolocations, and extends to personal information
such as email addresses, phone numbers, and financial information.\sidenote{\url{https://dimewiki.worldbank.org/wiki/De-identification}}
In many contexts this list may be more extensive --
for example, if you are working in a small environment,
someone's age and gender may be sufficient to identify them
even though these would not be considered PII in a larger context.
Therefore you will have to use your judgment in each case
to decide which pieces of information fall into this category.
This is \textit{usually} pretty straightforward outside of extremely specialized contexts.

\subsection{Handling personally-identifying information}

Raw data which contains PII \textit{must} be \textbf{encrypted}
during collection, storage, and transfer.
Most modern data collection software makes the first part straightforward.\sidenote{\url{https://dimewiki.worldbank.org/wiki/SurveyCTO_Form_Settings}}
However, secure storage and transfer are your responsibility.
One method for doing this is to keep raw data in a secure cloud location such as Amazon Web Services S3.
Another is to use an enterprise-grade solution such as Microsoft OneDrive; but these can be expensive.
\marginnote{Dropbox is not a secure storage location by default, and neither is GitHub.}
Depending on your specific security requirements,
it may be acceptable to create an encrypted version of the dataset
using software like VeraCrypt\sidenote{\url{https://www.veracrypt.fr/}}, and store that in Dropbox.
To securely transfer information,\marginnote{Email is \textit{never} a secure data transfer method.}
you can use the combination of Dropbox with encrypted data only,
or you can use another service such as WeTransfer\sidenote{\url{https://wetransfer.com}}.

In general, though, you shouldn't need to handle PII data very often.
Once data is securely collected and stored, the first thing you will generally do is \textbf{de-identify} it.
\index{De-identified data}
\marginnote{De-identified data should avoid, for example, you being sent back to every household to alert them that someone dropped all their personal information on a public bus and we don't know who has it.}
This simply means creating a copy of the data that contains no personally-identifiable information.
This data should be an exact copy of the raw data,
except it would be okay for it to be publicly released.\cite{matthews2011data}
Personally-identifiable information is most commonly used in field work, such as sampling and tracking,
and is rarely very useful in analysis. (The exception is geodata and this is an evolving field.)
This is very convenient for you, as you can simply leave the PII copy encrypted on disk as a backup,
and work primarily from the de-identifed dataset for almost all analytical tasks.

\subsection{Security and password management}

If you are handling data properly, it is going to end up password-protected in one way or another.
In a collaborative, long-term project, you need to be able to store and share these passwords.
\marginnote{You will also notice that your personal accounts contain an awful lot of sensitive information.}
Therefore you are going to have to get smart about password management and account security.
We recommend a service like LastPass.\sidenote{\url{https://www.lastpass.com}}
LastPass is free for personal features and cheap for sharing features.
It allows you to store an unlimited number of passwords, and you will have many.
You will have one (or more) for every website, for every project, for every encrypted dataset, and so on.
A service like LastPass allows you to store and share these passwords inside its own ecosystem,
since emailing passwords is just as vulnerable to theft or loss as writing them down.

LastPass also provides an app-based implementation for \textbf{two-factor authentication}
both for its own service and for other services you use (Gmail, Dropbox, Facebook, Amazon, etc.),
and we recommend you enable this feature for both your business and personal accounts.
Two-factor authentication (2FA) means you have to enter a second confirmation using your mobile device
to log into any of your password-protected services from new locations.
This is important not because anyone is up to anything nefarious,
but because random hacking is everywhere these days
and can arbitrarily damage, delete, or lock you out of your files in the worst case.

%------------------------------------------------

\section{Transparency in development research}

The empirical revolution in development research has not quite caused a crisis,
but it has certainly led to increased public scrutiny of the reliability of research.\cite{rogers_2017}
Three major components make up this scrutiny: credibility,\cite{ioannidis2017power} transparency,\cite{christensen2018transparency} and replicability.\cite{duvendack2017meant}
Development researchers take these concerns seriously.
This is because many development research projects are now purpose-built --
they cover specific topical questions, utilize unique data, and may use small samples.
On the one hand, this opens the door to working with the broader development community
to answer both specific programmatic questions as well as general research inquiries.
On the other, because the program and data researchers utilize today has never been seen by anyone else,
it is hard to verify that it was collected, handled, and analyzed appropriately.
Transparency is the most important way that researchers can avoid serious error,
and therefore is not a by-product but a core component of research output.

\subsection{Research credibility}

The credibility of research is primarily a function of design choices.
Is the research design sufficiently powered through its sampling and randomization?
Were the key research outcomes pre-specified or chosen ex-post?
How sensitive are the results to changes in specifications or definitions?
Tools such as \textbf{pre-analysis plans}\sidenote{\url{https://dimewiki.worldbank.org/wiki/Pre-Analysis_Plan}} are important to assuage these concerns for experimental evaluations,
\index{Pre-analysis plan}
but they may feel like ``golden handcuffs'' for other types of research.\cite{olken2015promises}
Regardless of whether or not a formal pre-analysis plan is utilized,
all experimental and observational studies should be pre-registered using the AEA database,\sidenote{\url{https://www.socialscienceregistry.org/}} the 3ie database,\sidenote{\url{http://ridie.3ieimpact.org/}}, or the OSF registry\sidenote{\url{https://osf.io/registries}} as appropriate.

Highly specific documentation is always helpful to establish credibility,
and will be expected in the supplemental materials to any publication.
The registries mentioned here will contribute to producing documentation,
but project documentation should always be an active and ongoing process,
not a one-time requirement or retrospective task.
New decisions are always being made as the plan begins contact with reality,
and there is nothing wrong with sensible adaptation so long as it is recorded and disclosed.
There are various software solutions for building these documentations over time.
The Open Science Framework\sidenote{\url{https://osf.io/}} provides one such solution,
with integrated file storage, version histories, and collaborative Wikipages.
GitHub\sidenote{\url{https://github.com}} provides a transparent task management platform
in addition to version histories and Wikipages, but is less effective for file storage.
The exact shape of this process should be agreed on by team members prior to project launch.

\subsection{Research transparency}

Transparent research will expose all of the processes involved in establishing credibility to the public.
This means that not only will readers be able to verify for themselves that the research was done well,
but they will be able to test assumptions and results that the researchers may not even have considered.
Well-structured research should make this as easy as possible for the reader to do.
\marginnote{We also hope to convince you that transparent research is easier for the researcher,
because it requires you to be organized in a labor-saving way.}
Etiquette is important here: nobody should go around trying to tear down the findings of others,
and neither should researchers guard or obfuscate their materials from public view.

\subsection{Research replicability}
