%------------------------------------------------

\begin{fullwidth}
Preparation for data work begins long before you collect any data.
In order to be prepared for fieldwork, you will need to know what you are getting into.
This means having a plan for the data sets you will need,
a plan for how those data sets will stay organized and linked,
and what identifying information each type of observation you'll collect.
This is called a \textbf{data map} for your project,
and it gives you and your team a sense of how information resources should be organized.
It's okay to update this once the project is underway --
the point is that everyone knows what the plan is.

At the same time, you also need to be technically prepared.
This means having a strong technical sense of your tools and workflow,
so that you do not spend a lot of time trying to figure out basic functions.
All of the tools we provide you here are designed to prepare you for collaboration and replication,
so that you can confidently manage tools and tasks on your computer.
We will try to provide free, open-source, and platform-agnostic tools wherever possible,
and provide more detailed instructions for those we are familiar.
However, most have a learning and adaptation process,
meaning you will become most comfortable with each tool
only by using it in real-world work.
\end{fullwidth}


%------------------------------------------------

\section{Preparing your digital workspace}

Being comfortable using your computer and having the tools you need in reach is key.
This section provides a brief introduction to key concepts and toolkits
that can help you take on the work that you will be primarily responsible for.
Some of these items skills may seem elementary,
but thinking about simple things from a workflow perspective
can help you make marginal improvements every day you work.

\subsection{Setting up your computer}

First things first: turn on your computer.
\marginnote{One reasonable setup is having your primary disk,
a local hard drive managed with a tool like Time Machine,
and a remote copy managed by a tool like Backblaze.
Dropbox files count only as local copies and never backups,
because others can alter it.}
Make sure you have fully updated the operating system,
that it is in good working order,
and that you have a password-protected login.
Then, make sure your computer is backed up.
Follow the 3-2-1 rule:
(3) copies of everything;
(2) different physical media;
(1) offsite storage.

There are a few things that can make your life much easier,
although some have a little expense associated with them.
\marginnote{This is the only time we'll suggest you spend money,
and it is totally okay if you or your organization cannot.
There are free options for almost everything except functional hardware.}
Mainly, make sure you have a \textit{good} computer.
Get at least 16GB of RAM and a 500MB or 1TB hard drive.
(Processor speeds matter less these days.)
Get a monitor with high-definition resolution.
Purchase authorized copies of software like Microsoft Office 365,
and pay for the individual tier of critical services like Dropbox,\sidenote{\url{https://www.dropbox.com}} Backblaze,\sidenote{\url{https://www.backblaze.com}} and LastPass.\sidenote{\url{https://www.lastpass.com}}
Get a decent email client (like Spark\sidenote{\url{https://sparkmailapp.com}} or Outlook),
a calendar that you like, the communication and note-taking tools you need,
a good music streaming service, and solid headphones with a mic.
None of these are essential to the work,
but they will make you a lot more comfortable doing it,
and being comfortable at your computer helps you stay happy and healthy.

\subsection{Preparing for collaboration and replication}

The first thing you need to have when working on your computer is a sense of where you are.
Find your \textbf{home folder}. On OSX, this will be a folder with your username.
On Windows, this will be something like ``This PC''. (It is never your desktop.)
Nearly everything we talk about will assume you are starting from here.
Ensure you know how to get the \textbf{absolute file path} for any given file.
On OSX this will be something like \path{/users/username/dropbox/project/...},
and on Windows, \path{C:/users/username/github/project/...}.

We will write filepaths such as \path{/Dropbox/project-title/DataWork/EncryptedData/}
using forward slashes, and mostly use only A-Z, dash, and underscore.
You should \textit{always} use forward slashes (\texttt{/}) in filepaths,
just like an internet address, and no matter how your computer writes them,
because the other type will cause your work to break many systems.
You can use spaces in names of non-technical files, but not technical ones.
Making the structure of your files part of your workflow is really important,
as is naming them correctly so you know what is where.\sidenote{\url{http://www2.stat.duke.edu/~rcs46/lectures_2015/01-markdown-git/slides/naming-slides/naming-slides.pdf}}


Next, you will need a personal working environment.
If you are working in R, RStudio is great.\sidenote{\url{https://www.rstudio.com}}
For Stata, we recommend the text editor Atom,\sidenote{\url{https://atom.io}}
which can open an entire targeted directory by writing \path{atom /path/to/directory/}
in the command line (Terminal or PowerShell), after copying it from the browser.
Opening the entire directory loads the whole tree view in the sidebar,
and gives you access to directory management actions.
You can start to manage your projects as a whole --
Atom is capable of sending code to Stata,\sidenote{\url{https://atom.io/packages/stata-exec}}
writing and building \LaTeX,\sidenote{\url{https://atom.io/packages/latex}}
and connecting directly with others to team code.\sidenote{\url{https://atom.io/packages/teletype}}
It is highly customizable, and since it is your personal environment,
there are lots of stylistic and functional options in extension packages
that you can use to make your work easier and more enjoyable.

You will also need a collaboration plan.
The two most common softwares in use are Dropbox and GitHub.\sidenote{\url{https://michaelstepner.com/blog/git-vs-dropbox/}}
Many organizations have an institutional subscription to one or both.
Dropbox will be preferred by most non-technical staff,
including project PIs and field staff.
It is preferable for sharing documents that are non-technical (like .docx),
and data that is non-sensitive (no personal information).
GitHub will be preferred for code,
and therefore most of what you are doing can be managed with Git.
Code is written in its native language,
and increasingly, outputs like reports\sidenote{\url{https://www.latex-project.org}}
and presentations\sidenote{\url{https://www.overleaf.com/learn/latex/Beamer}}
are being written in languages like \LaTeX.

Most importantly, talk to others in your team.
Figure out what datasets and code inputs are going to be created,
and agree on what is going to be stored where.
Agree on a folder structure: we provide
\texttt{iefolder}\sidenote{\url{hhttps://dimewiki.worldbank.org/wiki/iefolder}}
as a baseline guide for teams.
Agree on a \textbf{version control}\cite{blischak2016quick} workflow such as
Git Flow,\sidenote{\url{https://nvie.com/posts/a-successful-git-branching-model/}}


%------------------------------------------------


\section{Preparing a mental workflow}
