We want to thank all the people who helped us get here, especially
Arianna Legovini, for her leadership at DIME, unending support of our work,
and detailed comments on this book; and
Florence Kondylis, for her leadership in founding and growing DIME Analytics
and supporting this project from the very first.
We also thank the following members
of DIME Analytics for their contributions
to the ideas in this book and their help organizing them:
Roshni Khincha, Avnish Singh, Patricia Paskov, Radhika Kaul,
Mizuhiro Suzuki, Yifan Powers, and Maria Arnal Canudo.
This work has been financially supported by the United Kingdom Foreign,
Commonwealth \& Development Office (FCDO) through the
 DIME i2i Umbrella Facility for Impact Evaluation at the World Bank.

Our graditude to the many people who read and offered feedback as the book took shape:
%Alphabetical by last
Stephanie Annijas,
Maria Camila Ayala Guerrero,
Kaustubh Chahande,
Thomas Escande,
Aram Gassama,
Steven Glover,
Nausheen Khan,
Robert Norling,
Michael Orevba,
Caio Piza,
Francesco Raffaelli,
Daniel Rogger,
Ankriti Singh,
Ravi Somani,
and Leonardo Viotti.
Although they number far too many to name individually,
we also thank all the members of DIME and its teams across all the years
for the innovative work they have done, the lessons learned,
and the team spirit that makes our work so fruitful and rewarding.

This published version of the book has been revised repeatedly
since its internal release in June 2019,
with extensive feedback from readers and experts.
We would additionally like to thank Vincenzo di Maro
(Manager, DIME3) for his support throughout this process,
as well as peer reviewers
David McKenzie (Lead Economist, DECFP),
Holly Krambeck (Program Manager, DECAT),
Alaka Holla (Program Manager, HEDGE),
Jim Shen (Senior Manager, Abdul Latif Jameel Poverty Action Lab),
Federica di Battista (Trialling Lead, UK FCDO),
and Gabriel Vicente, Rajee Kanagavel, and Maksim Pecherskiy
from the Development Data Partnership team.

This book is a living product that is written and maintained publicly.
The code and edit history are at:
\url{https://github.com/worldbank/dime-data-handbook}.
You can get a PDF copy at:
\url{https://worldbank.github.com/dime-data-handbook}.
The website includes updated instructions
for providing feedback, as well as notes on updates to the content.
Whether you work with DIME, the World Bank,
or another organization or university,
we ask that you read the contents of this book critically.
We welcome feedback and corrections to improve the book.
Please visit
\url{https://worldbank.github.com/dime-data-handbook/feedback}
to provide feedback.
You can also email us at \url{dimeanalytics@worldbank.org},
and we will be very thankful.
We hope you enjoy \textit{Development Research in Practice}!
